---
title: "climate_dialy"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10, fig.height = 7
)
```

## Introduction
Interactive plot of the maximum, mean and minimum dialy temperatures for the study sites. Those data come from a grid of 500x500 m of resolution. 

```{r}
library(tidyverse)
library(dygraphs)
library(xts)
```

We selected the following sites: 

- NSH: Nevada P. sylvestris Healthy 
- NSD: Nevada P. sylvestris Decline
- BSD: Baza P. sylvestris Decline 
- ASH: Almijara P. sylvestris Healthy

```{r}
climate_grid <- read_csv("data/climate_rediam_dialy.csv") 

# selected sites
selected_sites <- c("NSH", "NSD", "BSD", "ASH1")


df_clima_dialy <- climate_grid |>
  dplyr::filter(name %in% selected_sites) |> 
  rename(code_site = name) |>
  mutate(date = as.Date(paste(year, month, day, sep = "-"), format = "%Y-%m-%d")) |> 
  mutate(code_site = case_when(
    code_site == "ASH1" ~ "ASH", 
    TRUE ~ code_site
  ))

tmax <- df_clima_dialy |> filter(var == "tmax") |> 
  dplyr::select(date, value, code_site) |> 
  pivot_wider(values_from = value, names_from = code_site)

tmed <- df_clima_dialy |> filter(var == "tmed") |> 
  dplyr::select(date, value, code_site) |> 
  pivot_wider(values_from = value, names_from = code_site)

tmin <- df_clima_dialy |> filter(var == "tmin") |> 
  dplyr::select(date, value, code_site) |> 
  pivot_wider(values_from = value, names_from = code_site)
```

### Dialy maximum temperatures

```{r}
tmax2019 <- tmax |> dplyr::filter(date > as.Date("2019-01-01", format = "%Y-%m-%d"))
tstmax <- xts(x=tmax2019, order.by = tmax2019$date)

min_values <- sapply(tmax2019[, -1], min, na.rm = TRUE)
max_values <- sapply(tmax2019[, -1], max, na.rm = TRUE)
y_range <- c(min(min_values) - 1, max(max_values) + 1)


cols_sites <- c(
  "ASH" = "#02BFFF",
  "BSD" = "#05FF01",
  "NSD" = "#FF00FF",
  "NSH" = "#FF00FF"
)


dygraph(tstmax) |> 
  dyRangeSelector() |> 
  dyAxis("y", label = "Dialy max temperature (ºC)", valueRange = y_range) |> 
  dyAxis("x", drawGrid = FALSE) |> 
  dySeries("ASH", color = "#02BFFF") |> 
  dySeries("BSD", color = "#05FF01") |> 
  dySeries("NSD", color = "#FF00FF", strokePattern = "dashed") |> 
  dySeries("NSH", color = "#FF00FF") 
  


```

### Dialy mean temperatures

```{r}
tmed2019 <- tmed |> dplyr::filter(date > as.Date("2019-01-01", format = "%Y-%m-%d"))
tstemd <- xts(x=tmed2019, order.by = tmed2019$date)

min_values <- sapply(tmed2019[, -1], min, na.rm = TRUE)
max_values <- sapply(tmed2019[, -1], max, na.rm = TRUE)
y_range <- c(min(min_values) - 1, max(max_values) + 1)


dygraph(tstemd) |> 
  dyRangeSelector() |> 
  dyAxis("y", label = "Dialy Mean temperature (ºC)", valueRange = y_range) |> 
  dyAxis("x", drawGrid = FALSE) |> 
  dySeries("ASH", color = "#02BFFF") |> 
  dySeries("BSD", color = "#05FF01") |> 
  dySeries("NSD", color = "#FF00FF", strokePattern = "dashed") |> 
  dySeries("NSH", color = "#FF00FF") 
  
```


### Dialy minimum temperatures

```{r}
tmin2019 <- tmin |> dplyr::filter(date > as.Date("2019-01-01", format = "%Y-%m-%d"))
tstmin <- xts(x=tmin2019, order.by = tmin2019$date)

min_values <- sapply(tmin2019[, -1], min, na.rm = TRUE)
max_values <- sapply(tmin2019[, -1], max, na.rm = TRUE)

# Calculate the y-axis range, considering all series
y_range <- c(min(min_values) - 1, max(max_values) + 1)

dygraph(tstmin) |> 
  dyRangeSelector() |> 
  dyAxis("y", label = "Dialy Mean temperature (ºC)", valueRange = y_range) |> 
  dyAxis("x", drawGrid = FALSE) |> 
  dySeries("ASH", color = "#02BFFF") |> 
  dySeries("BSD", color = "#05FF01") |> 
  dySeries("NSD", color = "#FF00FF", strokePattern = "dashed") |> 
  dySeries("NSH", color = "#FF00FF") 
  
```



```{r, eval=FALSE, echo=FALSE}

f <- list.files("data/climate_real/", full.names = TRUE) 
custom_locale <- locale(decimal_mark = ",")

read_custom <- function(x){ 
  o <- read_csv2(x, locale = custom_locale)
  result <- o |> 
    mutate(INDICATIVO = as.character(INDICATIVO)) |> 
    mutate(FECHA = as.Date(FECHA, format = "%d/%m/%Y"))
  return(result)
  }

all <- f |> map(~read_custom(.x)) |> reduce(bind_rows)

temp2018 <- all |> 
  dplyr::filter(FECHA > as.Date("2019-01-01", format = "%Y-%m-%d"))


```



