---
title: "Climate characterization"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10, fig.height = 7
)
```

```{r}
library(tidyverse)
library(ggpubr)
library(kableExtra)
library(ggpmisc)
library(data.table)

spei_data <- read_csv("data/spei_extracted.csv")
```



## Introduction 
- Assessment of the drought indices using SPEI Data since. 
- Two indices were used: SPEI-6 months and SPEI-12 months 
- Although we have the entire time series we will focus (the plots) on the period from 2010 onwards. 

### Time-series exploration
```{r}
plot_spei <- function(spei_data, spei_indices, title, sites) {
  spei_selected <- spei_data |>
    filter(spei %in% spei_indices) |>
    filter(site %in% sites) |> filter(!is.na(value)) |>
    mutate(
      sign = ifelse(value > 0, "pos", "neg"),
      variable = recode(spei,
                        "spei01" = "1-month",
                        "spei03" = "3-month",
                        "spei06" = "6-month",
                        "spei09" = "9-month",
                        "spei12" = "12-month",
                        "spei24" = "24-month",
                        "spei48" = "48-month"
      )
    )

  p <- ggplot(spei_selected, aes(x = date, y = value, fill = sign)) +
    geom_bar(stat = "identity") +
    facet_grid(site~variable) +
    scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
    theme_bw() +
    labs(
      title = title,
      x = "",
      y = "SPEI"
    ) +
    theme(
      strip.background = element_rect(fill = "white"),
      legend.position = "none"
    )

  return(p)
}
```

```{r}
sites <- c("BSD", "ASH1", "NSD", "NSH")

spei2010 <- spei_data |> 
  filter(date > as.Date("2010-01-01", format = "%Y-%m-%d"))

plot_spei(spei_data = spei2010,
                                 spei_indices = c("spei06"),
                                 sites = sites,
                                 title = "")
```

```{r}
plot_spei(spei_data = spei2010,
                                 spei_indices = c("spei12"),
                                 sites = sites,
                                 title = "")
```




### Compare the severe drought events. 
A severe drought event starts when SPEI falls below the threshold of −1.28 (Páscoa et al 2017; Spinoni et al. 2018). A drought event is considered only when SPEI values fall below that threshold for at least two consecutive months. For each drought event, we computed: 

- the duration as the number of consecutive months with the SPEI lower than a certain threshold
- the severity as the sum of the absolute SPEI values during the drought event
- the intensity and the Lowest SPEI refer to the mean and lowest value of SPEI, respectively, during the drought event.

We computed the severe drought events (below -1.28) by site and per each index (i.e. SPEI-06 and SPEI-12) since 1960. 

```{r}
droughtIndicators <- function(df, vname, threshold) {

  require(data.table)

  # Add a new column indicating if the value is below the threshold and the following month is also below the threshold
  out <- df |>
    mutate(is_drought = ifelse(
      .data[[vname]] < threshold & lead(.data[[vname]], default = .data[[vname]][n()]) < threshold, 1, 0
    )) |>
    mutate(date = lubridate::make_date(year, month))
  # Compute the drought duration of the events
  out2 <- out |>
    group_by(index_events = data.table::rleid(is_drought)) |>
    mutate(drought_duration = sum(is_drought)) |>
    as.data.frame()

  # Filter events with drought duration > 1
  out3 <- out2 |>
    filter(drought_duration > 1) |>
    as.data.frame()

  # Compute several indicators (drought assessments)
  da <- out3 |>
    group_by(index_events) |>
    summarise(
      d_duration = unique(drought_duration),
      d_intensity = mean(.data[[vname]], na.rm = TRUE),
      d_severity = sum(abs(.data[[vname]]), na.rm = TRUE),
      lowest_spei = min(.data[[vname]]),
      month_peak = month[which.min(.data[[vname]])],
      minyear = min(year),
      maxyear = max(year),
      rangeDate = paste(
        lubridate::month(min(date), label = TRUE), "-",
        (lubridate::month(max(date), label = TRUE))
      )
    ) |>
    as.data.frame()

  return(list(data = out2, drought_events = out3, drought_assessment = da))
}
```


```{r}
spei_mensual  <- spei_data |> 
  filter(spei %in% c("spei06", "spei12")) |> 
  filter(site %in% sites) |> 
  dplyr::select(site, spei, value, date) |> 
  mutate(year = lubridate::year(date), 
         month = lubridate::month(date)) |> 
  group_by(site, spei, year, month) |> 
  summarise(value = mean(value, na.rm = TRUE))


spei_events <- spei_mensual |> 
  group_by(spei, site) |> 
  nest() |> 
  mutate(drought_results = map(data, ~droughtIndicators(., "value", -1.28)$drought_assessmen)) %>%
  unnest(drought_results) |> 
  select(-data)

write_csv(spei_events, "data/spei_events.csv")
```

```{r}
spei_events |> 
  filter(spei == "spei06") |> 
  kbl(caption = "Drought periods by site (SPEI-06)") |> 
    kable_styling()
```

```{r}
spei_events |> 
  filter(spei == "spei12") |> 
  kbl(caption = "Drought periods by site (SPEI-12)") |> 
    kable_styling()
```

