---
title: "Climate characterization"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10, fig.height = 7
)
```

```{r}
library(tidyverse)
library(ggpubr)
library(kableExtra)
library(ggpmisc)
library(data.table)
library(patchwork)
library(cowplot)
```

```{r}
spei_data <- read_csv("data/spei_extracted.csv") |> 
  mutate(site = case_when(
    site == "ASH1" ~ "ASH", 
    TRUE ~ site
  )) |> 
  filter(date < as.Date("2022-06-01", format = "%Y-%m-%d")) 


spei_datam <- spei_data |> 
  mutate(date = lubridate::floor_date(date, unit = "month")) |> 
  group_by(site, spei, date) |> 
  summarize(value = mean(value, na.rm = TRUE))

```


## Introduction 
- Assessment of the drought indices using SPEI Data since. 
- Two indices were used: SPEI-6 months and SPEI-12 months. Different temporal interval could be computed. According to [Vicente-Serrano et al. 2010](10.1175/2009JCLI2909.1) short time scales are mainly related to soil-water content and river discharge in headwater areas, medium time scales are related to reservoir storages and discharge in the medium course of the rivers, and longtime scales are related to variations in groundwater storage.
- Although we have the entire time series we will focus (the plots) on the period from 2000 onwards. 
- We filtered data up to May-2022 included

## Time-series exploration
```{r}
plot_spei <- function(spei_data, spei_indices, title, sites, ...) {
  spei_selected <- spei_data |>
    filter(spei %in% spei_indices) |>
    filter(site %in% sites) |> filter(!is.na(value)) |>
    mutate(
      sign = ifelse(value > 0, "pos", "neg"),
      variable = recode(spei,
                        "spei01" = "1-month",
                        "spei03" = "3-month",
                        "spei06" = "6-month",
                        "spei09" = "9-month",
                        "spei12" = "12-month",
                        "spei24" = "24-month",
                        "spei48" = "48-month"
      )
    )

  p <- ggplot(spei_selected, aes(x = date, y = value, fill = sign, colour = sign)) +
    geom_bar(stat = "identity") +
    facet_grid(site~variable) +
    scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) + 
    scale_colour_manual(values = c("pos" = "darkblue", "neg" = "red")) +
    theme_bw() +
    labs(
      title = title,
      x = "",
      y = "SPEI"
    ) +
    theme(
      strip.background = element_rect(fill = "white"),
      legend.position = "none"
    )

  return(p)
}
```


```{r}
sites <- c("BSD", "ASH", "NSD", "NSH")

spei_selected <- spei_data |> 
  filter(date > as.Date("2000-01-01", format = "%Y-%m-%d")) 

spei_selectedm <- spei_datam |> 
  filter(date > as.Date("2000-01-01", format = "%Y-%m-%d")) 

treshold <- -1.28
```


### SPEI-6 
#### Weekly plot 

```{r}
plot_spei(spei_data = spei_selected,
                                 spei_indices = c("spei06"),
                                 sites = sites,
                                 title = "") + 
  geom_hline(yintercept = treshold, colour = "black", linetype = "dashed") + 
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  theme(panel.grid.minor.x = element_blank())

```

### Monhtly-plot
```{r}
plot_spei(spei_data = spei_selectedm,
                spei_indices = c("spei06"),
                sites = sites,
                title = "") +
  geom_hline(yintercept = treshold, colour = "black", linetype = "dashed") + 
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  theme(panel.grid.minor.x = element_blank())
```


### SPEI-12 
#### Weekly plot 
```{r}
plot_spei(spei_data = spei_selected,
                                 spei_indices = c("spei12"),
                                 sites = sites,
                                 title = "") +
  geom_hline(yintercept = treshold, colour = "black", linetype = "dashed") + 
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  theme(panel.grid.minor.x = element_blank())
```

### Monhtly-plot
```{r}
plot_spei(spei_data = spei_selectedm,
                spei_indices = c("spei12"),
                sites = sites,
                title = "") +
  geom_hline(yintercept = treshold, colour = "black", linetype = "dashed") + 
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  theme(panel.grid.minor.x = element_blank())
```


## Comparison of SPEI bewteen sites 

```{r}
spei06 <- spei_data |> 
  filter(spei == "spei06") |> 
  filter(date > as.Date("2000-01-01", format = "%Y-%m-%d")) |> 
  filter(site %in% sites) |> 
  na.omit() |> 
  dplyr::select(date, value, site) |> 
  pivot_wider(names_from = site, values_from = value) |> 
  dplyr::select(date, ASH:NSD)
```

### Bootstraped Pearson Correlation 

```{r}
custom_bci <- function(data, site1, site2, nrep){ 
  result <- confintr::ci_cor(data[c(site1,site2)], method = "spearman", type = "bootstrap", R = nrep)
  
  cor_value <- result$estimate[1] 
  ci <- result$interval
  vars <- dimnames(result$estimate)
  
  cor_df <- data.frame(
    v1 = vars[[1]], 
  v2 = vars[[2]], 
  correlation = cor_value,
  ci_lower = ci[1],
  ci_upper = ci[2]
)
  return(cor_df)
}
```

```{r}
set.seed(1234)
column_names <- c("ASH", "NSH", "BSD", "NSD")

# Create all possible pairs of column names
column_pairs <- combn(sites, 2, simplify = FALSE)

# Apply custom function using purrr::map_df
final_result <- purrr::map_df(column_pairs, ~ custom_bci(spei06, .[[1]], .[[2]], nrep = 5000))
```

```{r}
final_result |> 
  kbl(caption = "Boostraped Pearson correlation between series") |> 
    kable_styling()
```


```{r, echo=FALSE, eval=FALSE}
final_result |> 
  unite("cors", v1:v2, remove = FALSE, sep = " vs ") |> 
  ggplot(aes(x=cors, y=correlation)) + 
 
  geom_linerange(aes(ymin = ci_lower, ymax = ci_upper)) + 
  coord_flip() +
  geom_point() + 
  theme_minimal()
```


### Visualization

```{r}
# Create a list to store individual plots
plot_list <- list()
ylab <- "SPEI-06"
# Loop through each column pair
for (i in 1:length(column_pairs)) {
  column_pair <- column_pairs[[i]]
  
  # Select the relevant columns and pivot them longer
  data <- spei06 %>%
    select(date, column_pair) %>%
    pivot_longer(-date)
  
  # Create a ggplot line plot for the current column pair
  plot <- ggplot(data, aes(date, y=value, group=name, colour=name)) +
    geom_line() +
    labs(title = paste(column_pair[1], "-", column_pair[2]),
        y = ylab, x ="") +
    scale_color_manual(values=c("blue", "black")) +
    theme_minimal() +
    theme(
      legend.position = "top", 
      legend.title = element_blank()
    )
  
  # Add the plot to the list
  plot_list[[i]] <- plot
}

# Combine all individual plots using patchwork
combined_plot <- wrap_plots(plotlist = plot_list, ncol = 2)

# Display the combined plot
print(combined_plot)

```







## Compare the severe drought events. 
A severe drought event starts when SPEI falls below the threshold of −1.28 (Páscoa et al 2017; Spinoni et al. 2018). A drought event is considered only when SPEI values fall below that threshold for at least two consecutive months. For each drought event, we computed: 

- the duration as the number of consecutive months with the SPEI lower than a certain threshold
- the severity as the sum of the absolute SPEI values during the drought event
- the intensity and the Lowest SPEI refer to the mean and lowest value of SPEI, respectively, during the drought event.

We computed the severe drought events (below -1.28) by site and per each index (i.e. SPEI-06 and SPEI-12) since 1960. 

```{r}
droughtIndicators <- function(df, vname, threshold) {

  require(data.table)

  # Add a new column indicating if the value is below the threshold and the following month is also below the threshold
  out <- df |>
    mutate(is_drought = ifelse(
      .data[[vname]] < threshold & lead(.data[[vname]], default = .data[[vname]][n()]) < threshold, 1, 0
    )) |>
    mutate(date = lubridate::make_date(year, month))
  # Compute the drought duration of the events
  out2 <- out |>
    group_by(index_events = data.table::rleid(is_drought)) |>
    mutate(drought_duration = sum(is_drought)) |>
    as.data.frame()

  # Filter events with drought duration > 1
  out3 <- out2 |>
    filter(drought_duration > 1) |>
    as.data.frame()

  # Compute several indicators (drought assessments)
  da <- out3 |>
    group_by(index_events) |>
    summarise(
      d_duration = unique(drought_duration),
      d_intensity = mean(.data[[vname]], na.rm = TRUE),
      d_severity = sum(abs(.data[[vname]]), na.rm = TRUE),
      lowest_spei = min(.data[[vname]]),
      month_peak = month[which.min(.data[[vname]])],
      minyear = min(year),
      maxyear = max(year),
      rangeDate = paste(
        lubridate::month(min(date), label = TRUE), "-",
        (lubridate::month(max(date), label = TRUE))
      )
    ) |>
    as.data.frame()

  return(list(data = out2, drought_events = out3, drought_assessment = da))
}

```


```{r}
spei_mensual  <- spei_data |> 
  filter(spei %in% c("spei06", "spei12")) |> 
  filter(site %in% sites) |> 
  dplyr::select(site, spei, value, date) |> 
  mutate(year = lubridate::year(date), 
         month = lubridate::month(date)) |> 
  group_by(site, spei, year, month) |> 
  summarise(value = mean(value, na.rm = TRUE))


spei_events <- spei_mensual |> 
  group_by(spei, site) |> 
  nest() |> 
  mutate(drought_results = map(data, ~droughtIndicators(., "value", -1.28)$drought_assessmen)) %>%
  unnest(drought_results) |> 
  select(-data)

write_csv(spei_events, "data/spei_events.csv")

```

#### Drought events SPEI-06
```{r}
spei_events |> 
  filter(spei == "spei06") |> 
  kbl(caption = "Drought periods by site (SPEI-06)") |> 
    kable_styling()
```

##### Comparing drought events since 2000 

```{r}
d <- spei_events |> 
  separate(rangeDate, into = c("minmonth", "maxmonth"), remove = FALSE) |> 
  unite("start_date", minyear, minmonth, sep="-", remove = FALSE) |> 
  unite("end_date", maxyear, maxmonth,  sep="-", remove = FALSE) |> 
  mutate(start_date = as.Date(paste0(start_date, "-01"), format="%Y-%b-%d")) |> 
  mutate(end_date = as.Date(paste0(end_date, "-01"), format="%Y-%b-%d")) |> 
  rowwise() |> 
  mutate(middate = (start_date + ((end_date - start_date)/2))) |> 
  filter(start_date > as.Date("2000-01-01", format = "%Y-%m-%d")) 

```

```{r}
d06 <- d |> 
  filter(spei == "spei06") |> 
  ungroup() |> 
  dplyr::select(site, d_duration, d_intensity, d_severity, lowest_spei, month_peak, start_date, end_date, middate) 

plot_intensity <- d06 |> 
  ggplot(aes(y = site)) +
  geom_segment(aes(x = start_date, xend = end_date, yend = site, size=abs(d_intensity)), 
               colour = "red", lineend = "butt") +
  theme_minimal() + 
  labs(size = "Drougth's Intensity", x="")
  

plot_severity <- d06 |> 
  ggplot(aes(y = site)) +
  geom_point(aes(x = middate, y = site, size = d_severity)) + 
  theme_minimal() + 
  labs(size = "Drougth's Severity", x="")


plot_grid(plot_intensity, plot_severity, ncol = 1, align = "v") + 
          plot_annotation(title = "SPEI-06")
```

#### Drought events SPEI-12 
```{r}
spei_events |> 
  filter(spei == "spei12") |> 
  kbl(caption = "Drought periods by site (SPEI-12)") |> 
    kable_styling()
```

##### Comparing drought events since 2000 
```{r}
d12 <- d |> 
  filter(spei == "spei12") |> 
  ungroup() |> 
  dplyr::select(site, d_duration, d_intensity, d_severity, lowest_spei, month_peak, start_date, end_date, middate) 

plot_intensity <- d12 |> 
  ggplot(aes(y = site)) +
  geom_segment(aes(x = start_date, xend = end_date, yend = site, size=abs(d_intensity)), 
               colour = "red", lineend = "butt") +
  theme_minimal() + 
  labs(size = "Drougth's Intensity", x="")
  

plot_severity <- d12 |> 
  ggplot(aes(y = site)) +
  geom_point(aes(x = middate, y = site, size = d_severity)) + 
  theme_minimal() + 
  labs(size = "Drougth's Severity", x="")


plot_grid(plot_intensity, plot_severity, ncol = 1, align = "v") + 
          plot_annotation(title = "SPEI-12")
```


